<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{Technical reference documentation}
-->

Technical reference documentation
=======================================

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(out.extra='style="display:block; margin: auto"', fig.align="center")
```

Introduction
------------

This page gathers some more technical aspects of the **rClr** package, and is not meant to be read by most users.
One aim is to document a few processes such as the build process.

Build process
----------------------------

## .Rd file documentation
```{r, eval=FALSE}
library(rClrDevtools) # https://github.com/jmp75/rClr-devtools
roxyRclr()
```

## Vignettes
You should start from a new R session; if you have regenerated doc from Roxygen, there may be some surprises with the result of the `system.file` function.

```{r, eval=FALSE}
library(rClrDevtools) # https://github.com/jmp75/rClr-devtools
library(rClr)
pkgDir <- tryFindPkgDir()
# seem to need to move working directory. Oddities with knitr figure output otherwise
originalDir <- getwd()
docDir <- file.path(pkgDir, 'inst', 'doc')
setwd(docDir)
vignettesRclr()
```

```{cmd}
set R="c:\Program Files\R\R-3.1.0\bin\x64\R.exe"
cd F:\src\codeplex
set BuildConfiguration=Debug
rm -rf rClr.Rcheck
rm rClr_0.*.tar.gz
```

Note that a call to R CMD build will now build the vignettes too, so maybe vignettesRclr is redundant.

```{cmd}
%R% CMD build rClr > logbuild.txt 2>&1
%R% CMD check --as-cran rClr_0.6-1.tar.gz > logcheck.txt 2>&1
%R% CMD INSTALL rClr_0.6-1.tar.gz
```

Diagnosing
---------------

Have warnings, but by defaults could not see where, what. The following treats warnings as errors:
```{r, eval=FALSE}
options(warn=2)
test_package('rClr')
# restore:
options(warn=0)
```